\documentclass[thesis.tex]{subfiles}

\begin{document}
\chapter{Showcase of eIris Inductive}
\begin{itemize}
    \item Explain that we will show an example of the new iInductive system
    \item First we define the inductive and show the created lemma's and definitions
    \item Next we use the induction schema in our new tactic to proof a lemma
\end{itemize}
\section{Defining a new inductive}
\begin{itemize}
    \item Able to use standard \coq inductive syntax
    \item Show example
    \item Show fix-point and explain
    \item Show other Lemmas and explain as well
\end{itemize}
\begin{coqcode}
    EI.ind 
    Inductive is_list (q : Qp) : val → iProp :=
      | empty_is_list : is_list q NONEV
      | cons_is_list l v tl : 
        l ↦{#q} (v,tl) -∗ is_list q tl 
        -∗ is_list q (SOMEV #l).  
\end{coqcode}
\section{Proving a lemma using the inductive}
\begin{itemize}
    \item Explain Lemma we will prove
    \item Show steps of proof
\end{itemize}
\begin{coqcode}
    Lemma ind_test_1 (q q' : Qp) (v : val) :
        is_list q v ∗ is_list q' v ∗-∗ is_list (q+q') v.
    Proof.
        iSplit.
        - eiIntros "[Hq Hq']".
        iRevert "Hq'".
        eiInduction "Hq" as "[IH | (%l' & %v' & %tl' & Hl' & IH & %Hy)]"; eiIntros "Hq'".
        + iApply is_list_unfold_2.
            iLeft.
            iFrame.
        + simplify_eq.
            iApply is_list_unfold_2.
            iRight.
            iExists l', v', tl'.
            eiDestruct "Hq'" as "[%Hl | (%l'' & %v'' & %tl'' & Hl & Hilq' & %Hv)]"; simplify_eq.
            iCombine "Hl' Hl" as "Hl" gives %[_ ?]; simplify_eq.
            iFrame.
            iDestruct "IH" as "[IH _]".
            iSplitL.
            * iApply ("IH" with "[$]").
            * by iPureIntro.
        - eiIntros "Hi".
        eiInduction "Hi" as "[%Ha | (%l & %v' & %tl & [Hq Hq'] & [[Hiq Hiq'] _] & %Hy)]".
        + simplify_eq.
            iSplitL.
            * iApply is_list_unfold.
            iLeft.
            by iPureIntro.
            * iApply is_list_unfold.
            iLeft.
            by iPureIntro.
        + iSplitL "Hq Hiq".
            * iApply is_list_unfold.
            iRight.
            iExists l, v', tl.
            iFrame.
            by iPureIntro.
            * iApply is_list_unfold.
            iRight.
            iExists l, v', tl.
            iFrame.
            by iPureIntro.
    Qed.
\end{coqcode}
\end{document}